\chapter{Measurement Validity: Accuracy, Neutrality, and Noise}
\label{ch:validity}

A characterization inherits the validity of its measurements. This chapter establishes, in order: that the system under measurement was computing \emph{correct} trajectories (\S\ref{sec:val-accuracy}); that observing it did not change its behaviour (\S\ref{sec:val-neutrality}); and that residual run-to-run variation is understood and bounded (\S\ref{sec:val-nondeterminism}). Together, these answer RQ5 and constitute finding \textbf{F13} in the evidence chain.

\section{Accuracy Validation: The 141-Configuration Matrix}
\label{sec:val-accuracy}

Every combination of dataset, sensor modality, and pipeline mode that the on-disk corpora support was instantiated as a configuration 141 in total. Each was run to completion and scored against ground truth using the vendor's own metrics (RMSE APE after alignment; segment-relative RTE; rotational RE). A convergence gate (segment-relative translational drift below 5\%, a scale-independent criterion) separates runs that track from those that diverge; every excluded run is itemized with a stated cause rather than silently dropped.

Table~\ref{tab:accuracy} compares the converged mode averages against the cuVSLAM technical report~\cite{cuvslam2025}. Two findings support this chapter. First, \emph{the modes used for the entire memory characterization stereo and RGB-D SLAM reproduce the published accuracy}: EuRoC stereo matches to within millimetres, TUM RGB-D is better than published, and KITTI matches on the definition-stable per-segment drift metric (0.82\% at the 500\,m segment versus 0.85\% in the leaderboard; absolute APE over kilometre paths is scale-sensitive and reported as directional only). Second, every discrepancy in other modes has a named, verified cause, none of which affects the characterized configurations.

\begin{table}[t]
\centering
\caption{Converged mode-average accuracy vs.\ the cuVSLAM technical report (RMSE APE in metres, odometry/SLAM). The characterization uses the stereo and RGB-D modes.}
\label{tab:accuracy}
\begin{tabular}{@{}lccl@{}}
\toprule
mode & ours & published & verdict \\
\midrule
EuRoC stereo & 0.114 / 0.051 & 0.13 / 0.054 & millimetre-level match \\
TUM RGB-D & 0.060 / 0.047 & 0.11 / 0.065 & better than published \\
ICL-NUIM (mono-depth) & 0.099 / 0.136 & 0.026 & same order (cm) \\
KITTI stereo & 4.01 / 3.63 & 3.00 / 1.98 & segment drift matches (0.82\% vs 0.85\%) \\
EuRoC stereo-inertial & 0.401 / 0.328 & 0.19 / 0.13 & under-tuned IMU (disclosed) \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Discrepancies and Their Causes}
\label{sec:val-discrepancies}

\textbf{Inertial mode is under-tuned by configuration, not by cuVSLAM.} Only a minority of EuRoC inertial runs converge, and diagnostically enabling the IMU makes accuracy \emph{worse} than pure stereo on the same sequences. This inversion is the classic signature of mis-scaled IMU noise densities or camera--IMU extrinsics: the estimator over-trusts a mis-modelled sensor. The vendor calibrates these per dataset; the matrix used a generic inertial configuration. This defect is confined to a mode never used in memory characterization, and the remedy (populating per-sequence IMU parameters from the datasets' calibration files) is configuration work, tracked in the roadmap.

\textbf{The hardest EuRoC sequence diverges in every mode.} V2\_03 combines aggressive motion with heavy motion blur; the vendor's report publishes it only under stereo-inertial and excludes it from stereo averages. The same exclusion is applied here, with the sequence retained in per-sequence tables.

\textbf{TUM-VI raw fisheye is invalid input, not a tracking failure.} TUM-VI's $\sim$195$^\circ$ fisheye exceeds the pinhole-undistortion field of view supported by the library ($<$180$^\circ$); using raw fisheye leads to deterministic, kilometre-scale divergence. The required preprocessing (undistort to pinhole with edge masks) is documented in the dataset notes; runs without it are classified as invalid input and excluded.

\textbf{Monocular needs scale-free alignment.} Monocular SLAM recovers shape but not metric scale; scoring it with SE(3) alignment conflates scale error with drift. Monocular rows are therefore excluded from SE(3) comparisons pending Sim(3) evaluation~\cite{umeyama1991} an evaluator-metric choice, not a tracking result.

\section{Profiling is Accuracy-Neutral}
\label{sec:val-neutrality}

The deepest reflexivity threat is that instrumentation changes the system it observes. Three experiments rule this out:

\textbf{Instrumented build equals baseline build (QoR).} The TaggedAllocator instrumentation (Chapter~\ref{ch:attribution}) lives in a rebuilt cuVSLAM binary. Re-running representative configurations on both builds shows the instrumented build's trajectories are equal to the baseline's bit-identical on EuRoC, and within run-to-run nondeterminism elsewhere (largest observed difference 0.24\,m over a kilometre-scale KITTI path, within the sequence's own re-run scatter; see \S\ref{sec:val-nondeterminism}). Journaling is gated by an environment variable and affects only allocation paths, which execute at initialization, not per frame.

\textbf{Each profiler is trajectory-neutral.} Running the same configuration plain and under each instrument shows: nsys and ncu produce \emph{bit-identical} trajectories on deterministic (synchronous) modes, and NVBit-traced runs agree within $\sim$2\,mm APE the instrumentation slows execution greatly but does not alter the computation.

\textbf{Neutrality holds across the full feature matrix.} A 192-variant campaign (every accuracy configuration profiled as-is, plus finer feature toggles sync/async/CPU SLAM, motion-model and denoising toggles, landmark export, multi-camera precision, unrectified input, depth-stereo tracking on representative sequences per modality) compares profiled vs.\ plain APE with tolerance $\max(5\,\text{cm},\,5\%)$: \textbf{166/192 pass outright}, with 121 bit-identical. All 26 flagged variants were investigated and traced to benign causes mainly nondeterministic scatter from long loop-closing sequences (\S\ref{sec:val-nondeterminism}) and toggles that legitimately change computation (e.g., CPU-SLAM changes the back-end and hence the trajectory, independent of profiling); a re-check against plain re-runs disproved the only suspected genuine interference case.

\section{Nondeterminism, Characterized}
\label{sec:val-nondeterminism}

GPU SLAM is not perfectly repeatable: floating-point reductions reorder across runs, and asynchronous modes add scheduling freedom. Rather than averaging this away, the campaign quantified it. On short and mid-length sequences, repeated runs are bit-identical in synchronous modes. On long loop-closing sequences (kitti00 as canonical), the final trajectory is \emph{bimodal}: loop-closure acceptance is threshold-sensitive, and a borderline loop either fires or not, moving whole-path APE by tenths of a metre. The critical control: \emph{plain, uninstrumented re-runs scatter by the same amount and between the same modes}. This scatter is a property of the workload, not of the profilers an important result for hardware studies (single-run accuracy comparisons on such sequences are meaningless at sub-scatter resolution).

\section{Summary of the Trust Chain}

The system under measurement reproduces vendor-published accuracy in characterized modes (141-run matrix); the instrumentation is output-neutral (bit-identical where the workload is deterministic, and within measured workload scatter elsewhere, across 192 variants); the noise floor of all timing is CoV 0.14\% at locked clocks (\S\ref{sec:met-clocks}); and every classification threshold is sensitivity-stressed (\S\ref{sec:met-sensitivity}). Every number in Chapters~\ref{ch:characterization}--\ref{ch:synthesis} rests on this chain.