\chapter{Machine-Independent Locality Analysis}
\label{ch:locality}

Counters summarize; address traces prove. This chapter reports the
DAMOV-style locality analysis computed from NVBit per-warp address traces:
the metrics (\S\ref{sec:loc-metrics}), the front-end result
(\S\ref{sec:loc-frontend}), the loop-closure-scan result and the
self-correction that produced it (\S\ref{sec:loc-correction}), and the
session-scale growth measurement (\S\ref{sec:loc-growth}).

\section{Trace-side metrics}
\label{sec:loc-metrics}

From each kernel's global-space address stream, per launch:

\textbf{Footprint} --- the number of distinct 32-byte sectors touched,
$\times$32\,B: the exact working-set size, measured rather than inferred
from traffic.

\textbf{Reuse-distance CDF} --- for each access, the count of distinct
sectors touched since that sector's previous access (LRU stack
distance~\cite{mattson1970stack}); the CDF evaluated at capacities from
64\,KiB to 48\,MiB \emph{is} the predicted hit rate of any cache of that
size, independent of any real cache's policy details. A flat curve means no
cache size helps: the misses are compulsory.

\textbf{Coalescing, trace-exact} --- sectors touched per warp access
(1--32), plus active lanes per access, separating true scatter from
divergence-induced apparent scatter.

\textbf{Inter-launch overlap} --- the Jaccard similarity
$|A \cap B|/|A \cup B|$ between consecutive launches' footprints: does a
recurring kernel re-touch the same data (cacheable across launches) or
migrate?

One rule governs all of these: \emph{memory-space filtering}. Only
global-space instructions (\code{LDG}/\code{STG}) enter locality analysis;
shared-memory (\code{LDS}/\code{STS}) and local-memory
(\code{LDL}/\code{STL}, register spill) records are bucketed separately.
This rule was not a precaution added in advance---it was forced by the
measurement failure documented in \S\ref{sec:loc-correction}, and is now the
methodology's most exportable lesson.

\section{The front-end is cache-immune streaming}
\label{sec:loc-frontend}

For every per-frame front-end kernel (image cast, pyramid scaling, gradient
convolutions, LK tracking), the reuse-distance CDF is \textbf{flat from
64\,KiB to 48\,MiB}: the hit rate a 64\,KiB cache would achieve is the hit
rate a 48\,MiB cache would achieve, because all reuse is at tiny distances
(caught by any cache) and everything else is a first touch of streaming
pixel data (caught by none). Global-space accesses run at $\sim$4.0 sectors
per warp access, fully coalesced, at 32.0 active lanes.

The architectural reading is sharp: \emph{no realizable cache
helps this tier}---its misses are compulsory. Bigger caches are the wrong
axis entirely; the only remedies are moving fewer bytes (algorithmic) or
consuming the stream before it reaches DRAM (near-sensor placement,
\S\ref{sec:char-transfers}). An early counter-only reading had suggested one
convolution kernel was a scattered ``data-layout target'' (15.4 sectors per
access) with high reuse; space filtering dissolved both signals---they were
shared-memory tile traffic, on-chip and irrelevant to DRAM. After the
correction, the front-end is \emph{uniformly} stream-clean, and the correct
optimization target is on-chip, not layout.

\section{The loop-closure scan: a self-correction on record}
\label{sec:loc-correction}

The loop-closure scan kernel \stTrack{} produced the project's most
instructive measurement arc, reported here in full because its methodological
value exceeds that of a clean result.

\textbf{Round one: counters vs.\ trace.} Nsight Compute counters read the
kernel as a scattered gather---18--30 sectors per request. The first NVBit
trace analysis said the opposite: 2.1 sectors per warp access, 99.4\% of
accesses touching $\leq$4 sectors---apparently tightly coalesced streaming.
The trace being nominally ground truth, the project's interim report
concluded the counter metric was a proxy artifact---``the trace overturns
the counters.''

\textbf{Round two: the join exposes the confound.} When the attribution join
(Chapter~\ref{ch:attribution}) later matched trace addresses against live
allocations, 88--98\% of most kernels' trace records matched \emph{no
allocation at all}---impossible for real data traffic. The explanation: the
tracer records \emph{every} memory space, and for \stTrack{}, \textbf{92.6\%
of all recorded accesses were register-spill traffic} in local space---
which is \emph{perfectly coalesced by construction}
(\S\ref{sec:bg-mem})---drowning the actual data accesses in the blended
statistics.

\textbf{Round three: space-filtered re-derivation.} Filtering to
global space only and re-deriving every table: \stTrack{}'s \emph{data}
accesses are \textbf{23.4--30.0 sectors per warp access, with only
2.3--6.0\% of accesses touching $\leq$4 sectors}---a scattered gather,
\emph{exactly what the counters had said}. The interim conclusion was
withdrawn in a dated, in-place correction; the classifier's G2 label stands;
and the two instruments now agree. Table~\ref{tab:sttrack} juxtaposes the
blended and corrected views.

\begin{table}[t]
\centering
\caption{The \stTrack{} correction: blended-space analysis vs.\
space-filtered re-derivation (TUM room-scale / KITTI street-scale).}
\label{tab:sttrack}
\begin{tabular}{@{}lccc@{}}
\toprule
metric & blended (all spaces) & global-only & local-only (spill) \\
\midrule
sectors / warp access & 2.1 & 23.4 / 30.0 & 2.0 \\
accesses $\leq$4 sectors & 99.4\,\% & 6.0 / 2.3\,\% & 100\,\% \\
share of recorded accesses & 100\,\% & $\sim$5--7\,\% & $\sim$92.6\,\% \\
per-scan footprint (MB) & 0.467 / 1.096 & 0.464 / 1.093 & $\sim$0.003 \\
inter-launch Jaccard & 0.67 / 0.90 & 0.67 / 0.90 & 1.0 \\
\bottomrule
\end{tabular}
\end{table}

Three durable lessons. \emph{Methodological}: unfiltered GPU address traces
are dominated by spill and scratchpad records; any locality or attribution
analysis that does not filter by memory space is unsound---a warning
directly applicable to other groups using binary-instrumentation traces.
\emph{Evidential}: two independent instruments agreeing \emph{after} a
documented reconciliation is stronger support than either alone; the
disagreement was the pipeline catching its own error. \emph{Architectural}:
the kernel's DRAM volume and its DRAM \emph{pattern} have different owners
---volume is dominated by compiler spill (a register-file/occupancy trade,
fixable on-GPU), while the data-side pattern is a genuine scattered gather
over the keyframe descriptors (a scatter-PiM or layout question). Only the
attribution of Chapter~\ref{ch:attribution} could have separated them.

\section{Session-scale growth and migration}
\label{sec:loc-growth}

Two properties of the scan survive the correction untouched (the spill
window is only $\sim$3\,KB of the footprint). First, the per-scan working
set \emph{grows with the map}: 0.46\,MB per scan at room scale (TUM) to
1.09\,MB at street scale (KITTI). Second, it \emph{migrates}: consecutive
scans overlap with Jaccard 0.67 (room) to 0.90 (street)---10--33\% of the
touched set turns over between scans. Each individual scan is nearly
L2-resident; the union over a session---the entire keyframe database,
scanned incrementally, growing without bound with distance travelled---is
what no cache holds. This is the measured locality half of the in-storage
case; the allocation half (the database's GPU-resident state is a fixed
buffer while the growing store lives host-side) is established in
Chapter~\ref{ch:attribution}.
